Naturalness and Long-Lived Beyond the Standard Model Particles

### 2017-07-29 tex/draft.tex
fix ^M newlines
start adding figures, eqs

\usepackage[final]{fixme}
\fxsetup{layout=footnote}
\usepackage[left]{lineno}
\usepackage{amsmath,amsthm,amssymb,mathtools,graphicx,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref,siunitx,cancel}
\usepackage{fullpage}

@@

\section{The Standard Model}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing \fxnote{weak word} accuracy. The SM consists of \num{12} spin-$\frac{1}{2}$ particles, the fermions, that make up all observed matter; \num{12} spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces; and one fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are \num{8} gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are catagorized by which representation of each group they fit into. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distignuishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where Q is electric charge and $I_3$ is the third component of weak isospin.

@@

    SM parameters, such as particle masses\fxnote{what else is corrected. How?}, are corrected by interactions with virtual particles through loop diagrams. Figure\fxnote{make and reference diagrams}, for example, shows the potential corrections to the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{check. be able to work out example} on $\Lambda$. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

    \fxnote{Feynman diagrams go here}

@@

    As a fundamental scalar, the Higgs is doubly cursed. As explained in \fxnote{reference SM section}, $\mHiggs$ is quadratically sensitive to $\Lambda$, but that's not all: the Higgs field also lacks the chiral and gauge symmetries enjoyed by the fermions and gauge bosons. These symmetries act as the custodial symmetries mentioned in naturalness statement (3) in that they protect the fermion and gauge boson masses from large corrections. Because the SM chiral and gauge symmetries only apply to the massless theory, the SM symmetry group would be enlarged if fermions and gauge bosons were massless. That is, if the fermions and gauge bosons were massless at tree level, the chiral and gauge symmetries would cause their masses to be uncorrected to all orders in perturbation theory. As massive particles, however, the chiral and gauge symmetries instead guarantee that any corrections to their tree-level masses are proportional to their tree-level masses, preventing any large fine tuning between correction terms.

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to the apparently unnatural Higgs mass\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass. Arguments against naturalness often involve the anthropic principle, which claims that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than three spatial dimensions, which lowers the Planck mass itself. Finally, the dominant BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below.

\section{Supersymmetry}
    \fxnote{seems like I should mention that naturalness favors lighter sparticles}  
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions\fxnote{I need to be better at explaining exactly what this means}. In its simplest form, SUSY results in one new boson for every SM fermion, one new fermion for every SM boson, and one new Higgs doublet.\fxnote{make sure I understand the new Higgs doublet} The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with ``ino" tacked on to the end (e.g. Higgsino or wino). Collectively the superpartners to the SM particles are known as sparticles.\fxnote{cite martin}

    When calculating contributions from loop diagrams, one finds that fermion loops differ in sign from boson loops, which means that in SUSY every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. In fact, exact supersymmetry predicts that the bare Higgs mass is uncorrected to all orders in perturbation theory, as shown below.

    \fxnote{exact SUSY diagrams and correction term goes here}

@@

    \fxnote{broken SUSY correction term goes here}

@@

    Figure \fxnote{reference atlas SUSY limits} summarizes limits on SUSY models from ATLAS searches. CMS limits are generally similar\fxnote{make sure this is true}, but a recent plot was not available.

>>>>

\usepackage[final]{fixme}
\fxsetup{layout=footnote}
\usepackage[left]{lineno}
\usepackage[demo]{graphicx}
\usepackage{amsmath,amsthm,amssymb,mathtools,caption,subcaption,mathrsfs,cancel,multirow,enumitem,setspace,sectsty,cite,hyperref,siunitx,cancel}
\usepackage{fullpage}

@@

\section{The Standard Model} \label{SM}
    The standard model of particle physics (SM) describes all known particles and their non-gravitational interactions and is experimentally verified to astonishing \fxnote{weak word} accuracy. The SM consists of \num{12} spin-$\frac{1}{2}$ particles, the fermions, that make up all observed matter; \num{12} spin-1 particles, the gauge bosons, that communicate the electromagnetic, weak, and strong forces; and one fundamental scalar, the Higgs boson, which will be described in detail below.

    All SM particles transform according to the gauged symmetry group\fxnote{be good at describing the meaning of `gauged symmetry group'} $\SUthreecSUtwoLUoneY$, where $\SUthreec$ is associated with color charge, $\SUtwoL$ is associated with weak isospin, and $\UoneY$ is associated with weak hypercharge. The gauge bosons are associated with the generators of these groups, thus there are eight gluons associated with $\SUthreec$, three W bosons associated with $\SUtwoL$, and one B boson associated with $\UoneY$. The physical gauge bosons, which are linear combinations of the Ws and B, are the $\Wp$, $\Wm$, $\Z$, and photon ($\gamma$). The fermions are catagorized by which representation of each group they fit into\fxnote{make sure I'm good at explaining this}. For example, $\SUthreec$ distinguishes the quarks (which live in the fundamental representation) from the leptons (which are singlets), and $\SUtwoL$ distinguishes the left-handed fermions (doublets) from the right-handed fermions (singlets). Finally, weak hypercharge further distignuishes SM particles according to the Gell-Mann--Nishijima formula, $Y=Q-I_3$, where Q is electric charge and $I_3$ is the third component of weak isospin.

@@

    SM parameters, such as particle masses\fxnote{what else is corrected. How?}, are corrected by interactions with virtual particles through loop diagrams. Fig.~\ref{loop_diagrams}, for example, shows the potential corrections to the Higgs and electron masses. Because the Higgs is a fundamental scalar, its corrections depend quadratically on the cutoff scale, $\Lambda$, which is the largest energy scale for which the SM is valid. The fermions and gauge bosons, on the other hand, only depend logarithmically\fxnote{check. be able to work out example} on $\Lambda$. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the bare Higgs mass and the terms that correct it.

    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.8\textwidth]{images/loop_diagrams}
    \caption{Caption!\fxnote{cite something}}
    \label{loop_diagrams}
    \end{center} \end{figure}

@@

    As a fundamental scalar, the Higgs is doubly cursed. As explained in section \ref{SM}, $\mHiggs$ is quadratically sensitive to $\Lambda$, but that's not all: the Higgs field also lacks the chiral and gauge symmetries enjoyed by the fermions and gauge bosons. These symmetries act as the custodial symmetries mentioned in naturalness statement (3) in that they protect the fermion and gauge boson masses from large corrections. Because the SM chiral and gauge symmetries only apply to the massless theory, the SM symmetry group would be enlarged if fermions and gauge bosons were massless. That is, if the fermions and gauge bosons were massless at tree level, the chiral and gauge symmetries would cause their masses to be uncorrected to all orders in perturbation theory. As massive particles, however, the chiral and gauge symmetries instead guarantee that any corrections to their tree-level masses are proportional to their tree-level masses, preventing any large fine tuning between correction terms.\fxnote{wordy as hell}

    \fxnote{Clarify that not technically a problem. Nature might just be that way. Anthropic principle almost like an attempt to restore naturalness to a universe that appears unnatural, says Andrew} There are three main approaches to the apparently unnatural Higgs mass\fxnote{specify}: (1) argue that it was never really a problem in the first place, (2) reduce the SM cutoff scale, and (3) introduce a new symmetry to protect the Higgs mass. Arguments against naturalness often involve the anthropic principle, which posits that it should come as no surprise that humans live in a universe whose fundamental parameters are tuned to produce carbon-based life. Alternatively, some theories, such as large extra dimensions, lower the SM cutoff by proposing that gravity is spread across more than three spatial dimensions, which lowers the Planck mass itself. Finally, the dominant\fxnote{what do I mean by dominant?} BSM theory that aims to solve the naturalness problem by introducing a new symmetry is supersymmetry, which is explained below.

\section{Supersymmetry}
    \fxnote{seems like I should mention that naturalness favors lighter sparticles}  
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions\fxnote{I need to be better at explaining exactly what this means}. In its simplest form, SUSY results in one new boson for every SM fermion, one new fermion for every SM boson, and one new Higgs doublet.\fxnote{make sure I understand the new Higgs doublet} The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-$\frac{1}{2}$ superpartners to the SM bosons are given the name of their SM counterpart with ``ino" tacked on to the end (e.g. Higgsino or wino).\fxnote{describe charginos, neutralinos?} Collectively the superpartners to the SM particles are known as sparticles.\fxnote{cite martin}

    When calculating contributions from loop diagrams, one finds that fermion loops differ in sign from boson loops, which means that in SUSY every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. In fact, exact supersymmetry predicts that the bare Higgs mass is uncorrected to all orders in perturbation theory, as shown in Fig.~\ref{exact_susy_correction} below.

    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.8\textwidth]{images/exact_susy_correction}
    \caption{Caption!\fxnote{cite something}}
    \label{exact_susy_correction}
    \end{center} \end{figure}

@@

    \noindent \begin{equation}
    \delta \mHiggs = .
    \label{higgs_susy_correction}
    \end{equation}\fxnote{look up}

@@

    Fig.~\ref{atlas_susy_searches} summarizes limits on SUSY models from ATLAS searches. CMS limits are generally similar\fxnote{make sure this is true}, but a recent plot was not available.

    \fxnote{legible when printed?}  
    \noindent \begin{figure}[htbp] \begin{center}
    \includegraphics[width=0.8\textwidth]{images/atlas_susy_summary}
    \caption{Summary of current experimental limits on SUSY models from ATLAS searches.\fxnote{cite something}}
    \label{atlas_susy_summary}
    \end{center} \end{figure}
